Efficiently summarizing relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-19 treestats_paper.tex

% Fortunately, the genotype matrix is not the only way to encode genetic data:
% an alternative would be to take advantage of the relatedness amongst
% individuals and represent the data using trees.
Genotype matrices are massively redundant, however, and several compression methods
have been proposed~\citep{durbin2014efficient} \jk{TODO more refs}.
The fundamental reason for this redundancy is that samples share ancestry:
related individuals will tend to share the same state at a particular variant site.
Trees describe these ancestral
relationships~\citep{felsenstein2004inferring,semple2003phylogenetics},
and provide a natural and elegant way of encoding genetic data,
not only recording the history of a sample but also massively compressing
the genotype matrix~\citep{ane2005missing}. Until recently, however,
tree based approaches could not be used to encode data from sexually
reproducing organisms as recombination results in many trees along the
genome~\citep{griffiths1991two,griffiths1996ancestral,rasmussen2014genome},
and there was no efficient way of representing this information.
The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity)
is a recently introduced data structure that
encodes these genealogical trees along a genome concisely.
It was introduced in the context of coalescent
simulation~\citep{kelleher2016efficient}, leading to scalability increases of
several orders of magnitude over existing methods. The methods were subsequently extended
and refined for forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree}, with similarly large efficiency gains.
Recent work has shown that tree sequence algorithms can
also be used to massively increase the scalability of methods for inferring
genome-wide genealogies, and making it possible to infer trees for millions of
samples~\citep{kelleher2019inferring}.
The key to the remarkable efficiency of tree sequence
algorithms is the way that shared structure in adjacent trees along the genome is encoded.
As one looks across the genome, genealogies change at recombination breakpoints in ancestors.
However, nearby trees tend to share much of their structure,
and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
are often shared across relatively long distances,
which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
This redundancy can be exploited to store genome sequence and the
associated genealogical relationships very compactly, and has formed the basis for
efficient algorithms to calculate several population genetics
statistics~\citep{kelleher2016efficient}.
This paper generalizes the approach to a much broader class of statistics,
supporting arbitrary tree topologies and patterns of allelic variation.
% Want to emphasise generality here: people might assume we're only dealing
% with the biallelic case.

@@

The basic relationship between tree shape and summaries of genetic variation
is this: if mutations are neutral then the expected number that occur on a
given branch of a tree is proportional to the length of that branch.
This ``duality'' (we will use the term more precisely later) has been
used extensively in deriving properties of statistics in models of random mating
\citep[e.g.,][]{tajima1983evolutionary,tavare1984lineofdescent,fu1995statistical},
as has the fact that it applies regardless of the underlying demographic model
\jk{CITE Konrad/Nick here?}
\citep[e.g.,][]{gillespie1979evolutionary,hudson1983properties,slatkin1991inbreeding,
mcvean2002genealogical,ralph2019empirical}. We build on this central
insight to define a flexible and computationally efficient framework for
population genetic statistics on tree sequences.
Roughly speaking, to compute a statistic we will
propagate ``weights'' additively up each marginal tree,
summarize these weights on each branch with a ``summary function'',
and aggregate these summaries to produce statistics.
\autoref{fig:divergence_diagram} shows a simple example of these procedures
of propagating weights up a tree and summarizing them,
used to calculate mean sequence divergence over a short sequence spanned by a single tree.

@@

%% Check that math:
% n <- 10; z <- rnorm(10); z <- (z - mean(z))/sd(z); g <- rbinom(n, size=1, prob=1/2); p <- mean(g)
% c(cov(z,g), sum(z*g)/(n-1))
% c(cor(g,z), sum(g*z)/sqrt(n*(n-1)*p*(1-p)))
\begin{example}[Patterson's $f_4$] \label{ex:site_f4}
    Given four disjoint groups of samples, $S_1$, $S_2$, $S_3$, and $S_4$,
    Patterson's $f_4(S_1, S_2; S_3, S_4)$ statistic for an allele with frequency $p_i$ in group $S_i$
    is $(p_1 - p_2)(p_3 - p_4)$ \jk{CITATION NEEDED}.
    To rewrite this as a sum over alleles, note that
    $p_1 - p_2 = p_1 (1 - p_2) - (1 - p_1) p_2$,
    and so the statistic counts with positive value
    alleles that split $S_1$ and $S_3$ from $S_2$ and $S_4$,
    and negative value ones that split $S_1$ and $S_4$ from $S_2$ and $S_3$.
    Therefore, if as before we
    let $\iw_j = \bone_{S_j}$
    % let $\iw_{ju} = 1$ if $u \in S_j$
    (so that $\iw_{ju}$ tells us the number of samples in $S_j$ descended from $u$),
    and write $n_i$ for the number of samples in $S_i$, and
    \begin{align*}
        f(x_1, x_2, x_3, x_4)
        =
        \frac{x_1}{n_1}
        \left(1 - \frac{x_2}{n_2}\right)
        \frac{x_3}{n_3}
        \left(1 - \frac{x_4}{n_4}\right)
        -
        \left(1 - \frac{x_1}{n_1}\right)
        \frac{x_2}{n_2}
        \frac{x_3}{n_3}
        \left(1 - \frac{x_4}{n_4}\right),
    \end{align*}
    then $\site(\iw, f)_{[a,b)}$ is equal to Patterson's $f_4(S_1, S_2; S_3, S_4)$ statistic
    for the region of the genome between $a$ and $b$.
\end{example}

@@

We examined the contributions of mutation and demography to noise
in two simulations where population genetic (site) statistics
are important in making inferences:
detecting recent selective sweeps,
and detecting introgression.
Figure~\ref{fig:sweep_duality} shows windowed
diversity along the chromosome following a few selective sweeps.
The top two plots compare ``branch'' diversity --- i.e., as computed only with tree shape ---
to ``site'' diversity computed from sequence generated by 20 independent assignments of mutations to the same tree sequence,
with mutation rates $10^{-9}$ and $10^{-8}$, respectively.
We see that as the mutation rate increases, the signal of decrease in diversity around swept loci becomes more clear,
and approaches ``branch'' diversity.
These were computed using the entire population of 1,000 individuals;
how does sampling variance contribute?
Not much, it turns out --- the bottom plot shows both site and branch diversity
computed from 20 nonoverlapping groups of 100 samples.
Neither site or branch diversity vary much between these samples,
implying that the subsample gives us a good estimate of the whole-population values of each.
However, as we see in the top figure,
whole-population site diversity is itself only a quite noisy estimator of branch diversity.
The same things are shown in Supplementary Figure \ref{sfig:sweep_duality_10K}
for a simulation with 10,000 individuals, which shows similar patterns.
Only the last of these plots is possible to directly observe in real data:
in the top two plots,
the spread of independent replicates of mutational noise (black lines)
about their expectation based on the tree sequence (red line)
is unobservable, although estimable.

@@

As another example, we simulated an ``admixture'' scenario:
a first population of size $N=1000$ splits into two of equal size,
then after $N$ generations, the second population splits,
after another $N$ generations, the third population splits again,
and for the final $N$ generations populations 2 and 3
have per-capita migration rates of $4/N$ to each other.
We expect a significantly negative $f_4(1,2;3,4)$ in this situation,
which we indeed find
(the genome-wide mean Branch $f_4$ is around -700,
as is the Site $f_4$ divided by mutation rate;
see Supplemental Figure \ref{fig:introf4} for a plot along the genome).
Using various sample sizes, mutation rates, and window sizes, we then
calculated this $f_4$ statistic in windows along a 100Mb genome,
and show the standard deviation of both ``site'' and ``branch'' statistics across windows
in Figure \ref{fig:introf4sd}.
Since the genome is uniform (no selection or variation in recombination or mutation),
this standard deviation is a measure of noisiness.
As expected, site statistics are noisier than branch statistics,
by a factor that depends mostly on the ratio of mutation to recombination rates.
The results suggest that branch statistics would provide substantially better resolution
at small scales along the genome, especially if mutation rate is lower than recombination rate.
However, in practice imperfect estimation of the tree sequence
would introduce additional noise,
so it remains to be seen if the improvement could be made in practice.

@@

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them at
scale~\citep{harris2019database,kelleher2019inferring,speidel2019method}.
In Figure~\ref{fig:sweep_duality}, we showed that branch and site statistics
matched well in simulated data. However, these simulations make many
simplifying assumptions, and, moreover,
the underlying tree topologies and branch lengths are exactly
correct (which inference methods can only hope to approximate).
To evaluate the correspondence between site and branch statistics
in tree sequences inferred from real data,
we calculated statistics for chromosome 20
of the 1000 Genomes dataset~\citep{1k2015global} using dated trees estimated
by Relate~\citep{speidel2019method}.
Specifically, we calculated diversity using the site statistic
described in Example~\ref{ex:site_diversity},
and compared this to the dual branch statistic (Example~\ref{ex:branch_diversity})
in 1Mb windows in each of the five continental groupings.
All calculations were done only using the portions of the chromosome
passing the 1000 Genomes Phase 1 ``strict'' mask for sequencing accessibility.
% from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
As shown in Figure~\ref{fig:chr20}, Branch and Site diversity along the genome are very similar.
Is this an indication that Relate is doing a good job at inferring trees,
or just a byproduct of the inference method (which, naturally, estimates branch lengths to be proportional to numbers of mutations)?
Answering this question requires a deeper understanding of the processes that shape
genetic diversity along the genome.

@@

In practice, we do not expect branch and site diversity to agree exactly,
because of local differences in mutation rate and intensities of selected mutations.
For instance, regions in which a high proportion of potential mutations are deleterious
are expected to have a lower diversity for two reasons:
first, the deleterious mutations themselves are less likely to be found,
and second, the indirect effect of deleterious mutations on nearby sites reduces typical tree height
and thus diversity at even neutral, linked sites \citep{hudson1994levels,charlesworth1997effects}.
The first effect would cause site and branch statistics to differ,
because it effectively reduces the mutation rate in the region
and violates the assumption of independence of mutations given the trees.
However, the second effect does \emph{not} affect the correspondence between site and branch statistics,
since it is mediated by tree shape.
It is generally unknown how much mutation rate varies along the genome
\citep[but see][]{supek2015differential}
or how dense are targets of selection \citep{leffler2012revisiting}.
For this reason, it is very interesting to see how close a correspondence between site and branch
statistics it is \emph{possible} to obtain ---
it is tempting to say that the best tree sequence would obtain
as tight a match between site and branch statistics as possible,
with remaining variation explained by direct selection and/or mutation rate variation.

@@

%%%%%%%%%%%%%%%%%%%%%
\subsection*{Algorithms}
In this section we describe and analyze the algorithm used
to compute Branch statistics. This is a generalization
of the algorithm used to maintain the number of samples from a
given set that derive from each node in a tree
sequence \cite[Algorithm L]{kelleher2016efficient}. The central
problem shared across site, branch and node statistics
is to efficiently maintain the ``state'' $x_u$ (where $x_u$ is the
sum of the weights for all samples descending
from, and including, node $u$; see Figure~\ref{fig:divergence_diagram}).
As we move along the sequence, the trees change when branches are inserted and removed.
By carefully ordering these insertions and removals,
we ensure that this state can be correctly maintained
using only small adjustments for each tree.
Computing the statistics is then a straightforward matter of applying the summary function $f$
to the node states and aggregating the results appropriately for the
particular statistics mode and windowing options.

@@

\begin{taocpalg}{B}{General branch statistics}
{Given a list of $n$ sample nodes $S$, corresponding weights $w$ and
summary function $f: \R \rightarrow \R$
compute the span-normalized statistic $\sigma$
over a tree sequence with sequence length $L$
defined by the node and edge tables $\Nt$ and $\Et$. We assume
that the index vectors $\indexin$ and $\indexout$ have been precomputed.
}

@@

Algorithm \algref{B} begins by setting the initial state for $\pi$, $\beta$,
$x$ and $F$ for each node in the tree sequence, and then sets the values
of $x_u$ and $F_u$ for each of the samples $u$ in $S$.
(The initial state is the ``empty forest'', where no nodes are connected to any others.)
Afterwards, we set our ``running sum'' $s$ and output statistic $\sigma$ to zero, along
with the ``tree left'' variable $t_l$. The $j$ and $k$ variables are
used to keep track of our position in the edge insertion and edge
removal indexes, respectively. After completing initialization in
\algref{B1}, we enter the main tree loop in \algref{B2} which is
run once for each tree in the sequence. As we are processing each
tree, we keep track of the edges that need to be processed with
the $t_l$ variable, which stores the left-most genome coordinate
in the current tree. The first thing we do for a new tree is to
remove any edges that were in the previous tree and are not in the
current tree. These must be the edges in which the right coordinate
is equal to $t_l$, and so \algref{B3} loops over these edges using
the edge removal index $\indexout$. Step \algref{B4} then removes
the branch $u \mapsto v$ corresponding to a single edge, by
subtracting its contribution to the running sum $s$, setting the parent of
$u$ to $-1$ and its branch length to $0$. As shown in
Figure~\ref{fig:algorithm_example}B, removing an edge will result
in the subtree rooted at $u$ become disconnected from the rest of the
tree. Step \algref{B5} ensures that the state of the rest of the tree
is correctly maintained by propagating the loss of the state $x_u$
up the tree from $v$. Thus, for each node $v$ that was a
direct ancestor of $u$ in the previous tree, we first subtract
its contribution from the running sum $s$ and remove the contribution
of $x_u$ from its state. Since the state of $v$ has changed, we must
recompute the value of our summary cache $F_v$, and then finally
add the new contribution of $v$ to $s$. (For example, note the changes in $x$
and $F$ for nodes 7 and 8 in Figure~\ref{fig:algorithm_example}B.)

@@

To analyze Algorithm~\algref{B} we will assume that the tree sequence has
been simulated under the standard coalescent with a sample of size
$n$ and population scaled recombination rate $\rho = 4 N_e r$,
where $r$ is the mean number of recombinations per chromosome per unit of time.
Under these conditions, there are $O(n + \rho \log n)$ edges in the tree sequence
\citep{kelleher2016efficient}.  Clearly, each edge is examined
exactly once in steps \algref{B6} and \algref{B7} in order to create
the trees, and at most once in steps \algref{B3} and \algref{B4} (we do
not remove the edges for the final tree). Additionally, each edge that
we insert or remove after the initial building of the tree
may incur the cost of traversing up the tree as
far as the root in steps \algref{B5} and \algref{B8}.
As coalescent genealogies are asymptotically
balanced \citep{li2013coalescent}, the expected number of nodes
on the path to root is $\log_2 n$.
Therefore, the overall running time
of Algorithms~\algref{B} is of order $n + \rho \log(n) \log_2(n)$, which is
\begin{equation}\label{eqn:alg_complexity}
    O\left(n + \rho \log(n)^2 \right).
\end{equation}
Computing a site statistic will have the same complexity,
as the same internal state is maintained,
and the number of segregating sites is proportional to the number of edges
(with constant equal to the ratio of mutation to recombination rate).
The asymptotic analysis here assumes that we traverse upwards to root
for each edge, but this is not the case. Edges are removed in
oldest-first order, guaranteeing that the deepest node any subtree being
modified is processed first. Therefore, this subtree will be disconnected from the
main tree, ensuring that we traverse upwards all the way to root only once.
Similarly, edges are inserted in youngest-first order, ensuring that subtrees are only
reconnected to the main tree when they are fully built.

@@

To determine how well our theoretical model of time complexity predicts the
running time in \tskit{}'s implementation of the site statistics algorithm,
we fitted a model based on Equation~\eqref{eqn:alg_complexity}.
\autoref{fig:stats_performance} shows that theoretical predictions match
the observed running time very well. As the overall complexity is
$O(n + \rho \log(n)^2)$, for sufficiently large $n$ the initial term
(representing the cost of building the first tree) will come to dominate.
In this case, we can see this happening at around $n = 10^5$, where there
is a noticeable uptick in the time required per variant. For
longer genomes (i.e., larger values of $\rho$), this cost is amortized over more trees
and is less apparent.

@@

It should be noted here that this example based on simulated data
represents a best-case scenario in terms of the performance advantages
of \tskit{} over the matrix-based methods. For real data, inferred
tree sequences currently contain substantially more edges than we
would expect based on simple neutral theory~\citep{kelleher2019inferring},
and therefore computing statistics will not be as efficient as
for an equivalent simulation. Tree sequence inference
methods are in their infancy, however, and it is likely that
as they improve the number of edges required to encode data will
be reduced. For data simulated natively in tree sequence format
via packages such as \texttt{msprime}~\citep{kelleher2016efficient},
\texttt{SLiM}~\citep{haller2018tree,haller2019slim} and
\texttt{fwdpy11}~\citep{thornton2014c++}, the advice is straightforward.
Computing statistics using the algorithms in \tskit{}
will always be more efficient than decoding the genotype
matrix, importing it into another package and computing from the matrix.

@@

% visualization and time decomposition adding a new dimension
Genomic data are naturally distributed along two axes: along the genome, and across geography.
The tree sequence extends this to a third axis: time.
A great many methods in population genetics aim to describe aspects of history,
and accurate (or at least unbiased) tree sequence inference
may shift the focus of these problems from inference to descriptive analysis.
The methods developed here distribute the contribution to various statistics across each tree,
and so could also be used to summarize contributions to various statistics across time.
This could provide, for instance, the time distribution of mutations or branches
contributing positively or negatively to $f_4$ statistics of introgression,
enabling historical interpretations of these signals.
% This additional mode of reporting statistics is an important subject for future work.
% moving past a few summaries in discrete groups to a genome/time/space view of relatedness
The computational toolbox of population genetics
is still mostly composed of statistics designed for analysis of a handful of loci
in a small number of discrete, mostly separated populations.
Both our data and our understanding of the world have moved beyond this.
We hope that the tools developed here will help make it possible
to visualize and analyze genetic variation and genealogical relatedness
along the genome, across space, and through time.

>>>>

% Fortunately, the genotype matrix is not the only way to encode genetic data:
% an alternative would be to take advantage of the relatedness amongst
% individuals and represent the data using trees.
Genotype matrices are massively redundant, however, and several specialised
compression methods
%JK Note for later: I think there's an earlier citation I'm missing here, but
% I can't find it right now. Look it up when doing later corrections.
have been proposed~\citep{durbin2014efficient,layer2016efficient,danek2018gtc,lin2019sparse}.
The fundamental reason for this redundancy is that samples share ancestry:
related individuals will tend to share the same state at a particular variant site.
Trees describe these ancestral
relationships~\citep{felsenstein2004inferring,semple2003phylogenetics},
and provide a natural and elegant way of encoding genetic data,
not only recording the history of a sample but also massively compressing
the genotype matrix~\citep{ane2005missing}. Until recently, however,
tree based approaches could not be used to encode data from sexually
reproducing organisms as recombination results in many trees along the
genome~\citep{griffiths1991two,griffiths1996ancestral,rasmussen2014genome},
and there was no efficient way of representing this information.
The \emph{succinct tree sequence} (or \emph{tree sequence}, for brevity)
is a recently introduced data structure that
encodes these genealogical trees along a genome concisely.
It was introduced in the context of coalescent
simulation~\citep{kelleher2016efficient}, leading to scalability increases of
several orders of magnitude over existing methods. The methods were subsequently extended
and refined for forwards-time
simulations~\citep{kelleher2018efficient,haller2018tree}, with similarly large efficiency gains.
Recent work has shown that tree sequence algorithms can
also be used to massively increase the scalability of methods for inferring
genome-wide genealogies, and making it possible to infer trees for millions of
samples~\citep{kelleher2019inferring}.
The key to the remarkable efficiency of tree sequence
algorithms is the way that shared structure in adjacent trees along the genome is encoded.
As one looks across the genome, genealogies change at recombination breakpoints in ancestors.
However, nearby trees tend to share much of their structure,
and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
are often shared across relatively long distances,
which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
This redundancy can be exploited to store genome sequence and the
associated genealogical relationships very compactly, and has formed the basis for
efficient algorithms to calculate several population genetics
statistics~\citep{kelleher2016efficient}.
This paper generalizes the approach to a much broader class of statistics,
supporting arbitrary tree topologies and patterns of allelic variation.
% Want to emphasise generality here: people might assume we're only dealing
% with the biallelic case.

@@

The basic relationship between tree shape and summaries of genetic variation
is this: if mutations are neutral then the expected number that occur on a
given branch of a tree is proportional to the length of that branch.
This ``duality'' (we will use the term more precisely later) has been
used extensively in deriving properties of statistics in models of random mating
\citep[e.g.,][]{tajima1983evolutionary,tavare1984lineofdescent,fu1995statistical},
as has the fact that it applies regardless of the underlying demographic model
\citep[e.g.,][]{gillespie1979evolutionary,hudson1983properties,slatkin1991inbreeding,
mcvean2002genealogical,lohse2016efficient,ralph2019empirical}. We build on this central
insight to define a flexible and computationally efficient framework for
population genetic statistics on tree sequences.
Roughly speaking, to compute a statistic we will
propagate ``weights'' additively up each marginal tree,
summarize these weights on each branch with a ``summary function'',
and aggregate these summaries to produce statistics.
\autoref{fig:divergence_diagram} shows a simple example of these procedures
of propagating weights up a tree and summarizing them,
used to calculate mean sequence divergence over a short sequence spanned by a single tree.

@@

%% Check that math:
% n <- 10; z <- rnorm(10); z <- (z - mean(z))/sd(z); g <- rbinom(n, size=1, prob=1/2); p <- mean(g)
% c(cov(z,g), sum(z*g)/(n-1))
% c(cor(g,z), sum(g*z)/sqrt(n*(n-1)*p*(1-p)))
\begin{example}[Patterson's $f_4$] \label{ex:site_f4}
    Given four disjoint groups of samples, $S_1$, $S_2$, $S_3$, and $S_4$,
    Patterson's $f_4(S_1, S_2; S_3, S_4)$
    statistic~\citep{reich2009reconstructing,patterson2012ancient}
    for an allele with frequency $p_i$ in group $S_i$
    is $(p_1 - p_2)(p_3 - p_4)$. To rewrite this as a sum over alleles, note that
    $p_1 - p_2 = p_1 (1 - p_2) - (1 - p_1) p_2$,
    and so the statistic counts with positive value
    alleles that split $S_1$ and $S_3$ from $S_2$ and $S_4$,
    and negative value ones that split $S_1$ and $S_4$ from $S_2$ and $S_3$.
    Therefore, if as before we
    let $\iw_j = \bone_{S_j}$
    % let $\iw_{ju} = 1$ if $u \in S_j$
    (so that $\iw_{ju}$ tells us the number of samples in $S_j$ descended from $u$),
    and write $n_i$ for the number of samples in $S_i$, and
    \begin{align*}
        f(x_1, x_2, x_3, x_4)
        =
        \frac{x_1}{n_1}
        \left(1 - \frac{x_2}{n_2}\right)
        \frac{x_3}{n_3}
        \left(1 - \frac{x_4}{n_4}\right)
        -
        \left(1 - \frac{x_1}{n_1}\right)
        \frac{x_2}{n_2}
        \frac{x_3}{n_3}
        \left(1 - \frac{x_4}{n_4}\right),
    \end{align*}
    then $\site(\iw, f)_{[a,b)}$ is equal to Patterson's $f_4(S_1, S_2; S_3, S_4)$ statistic
    for the region of the genome between $a$ and $b$.
\end{example}

@@

We examined the contributions of mutation and demography to noise
in two simulations where population genetic (site) statistics
are important in making inferences:
detecting recent selective sweeps,
and detecting introgression.
Figure~\ref{fig:sweep_duality} shows windowed
diversity along the chromosome following a few selective sweeps.
The top two plots compare Branch diversity --- i.e., as computed only with tree shape ---
to Site diversity computed from sequence generated by 20 independent assignments of mutations to the same tree sequence,
with mutation rates $10^{-9}$ and $10^{-8}$, respectively.
We see that as the mutation rate increases, the signal of decrease in diversity around swept loci becomes more clear,
and approaches Branch diversity.
These were computed using the entire population of 1,000 individuals;
how does sampling variance contribute?
Not much, it turns out --- the bottom plot shows both site and branch diversity
computed from 20 nonoverlapping groups of 100 samples.
Neither site or branch diversity vary much between these samples,
implying that the subsample gives us a good estimate of the whole-population values of each.
However, as we see in the top figure,
whole-population site diversity is itself only a quite noisy estimator of branch diversity.
The same things are shown in Supplementary Figure \ref{sfig:sweep_duality_10K}
for a simulation with 10,000 individuals, which shows similar patterns.
Only the last of these plots is possible to directly observe in real data:
in the top two plots,
the spread of independent replicates of mutational noise (black lines)
about their expectation based on the tree sequence (red line)
is unobservable, although estimable.

@@

As another example, we simulated an ``admixture'' scenario:
a first population of size $N=1000$ splits into two of equal size,
then after $N$ generations, the second population splits,
after another $N$ generations, the third population splits again,
and for the final $N$ generations populations 2 and 3
have per-capita migration rates of $4/N$ to each other.
We expect a significantly negative $f_4(1,2;3,4)$ in this situation,
which we indeed find
(the genome-wide mean Branch $f_4$ is around -700,
as is the Site $f_4$ divided by mutation rate;
see Supplemental Figure~\ref{fig:introf4} for a plot along the genome).
Using various sample sizes, mutation rates, and window sizes, we then
calculated this $f_4$ statistic in windows along a 100Mb genome,
and show the standard deviation of both Site and Branch statistics across windows
in Figure~\ref{fig:introf4sd}.
Since the genome is uniform (no selection or variation in recombination or mutation),
this standard deviation is a measure of noisiness.
As expected, site statistics are noisier than branch statistics,
by a factor that depends mostly on the ratio of mutation to recombination rates.
The results suggest that branch statistics would provide substantially better resolution
at small scales along the genome, especially if mutation rate is lower than recombination rate.
However, in practice imperfect estimation of the tree sequence
would introduce additional noise,
so it remains to be seen if the improvement could be made in practice.

@@

We do not know the true genealogies underlying real data,
but recent methods are available to estimate them at
scale~\citep{kelleher2019inferring,speidel2019method}.
In Figure~\ref{fig:sweep_duality}, we showed that branch and site statistics
matched well in simulated data. However, these simulations make many
simplifying assumptions, and, moreover,
the underlying tree topologies and branch lengths are exactly
correct (which inference methods can only hope to approximate).
To evaluate the correspondence between site and branch statistics
in tree sequences inferred from real data,
we calculated statistics for chromosome 20
of the 1000 Genomes dataset~\citep{1k2015global} using dated trees estimated
by Relate~\citep{speidel2019method}.
Specifically, we calculated diversity using the site statistic
described in Example~\ref{ex:site_diversity},
and compared this to the dual branch statistic (Example~\ref{ex:branch_diversity})
in 1Mb windows in each of the five continental groupings.
All calculations were done only using the portions of the chromosome
passing the 1000 Genomes Phase 1 ``strict'' mask for sequencing accessibility.
% from ftp://ftp.1000genomes.ebi.ac.uk/vol1/ftp/phase1/analysis_results/supporting/accessible_genome_masks/README.20120515_phase1_pop_gen_mask
As shown in Figure~\ref{fig:chr20}, Branch and Site diversity along the genome are very similar.
Is this an indication that Relate is doing a good job at inferring trees,
or just a byproduct of the inference method (which, naturally, estimates branch lengths to be proportional to numbers of mutations)?
Answering this question requires a deeper understanding of the processes that shape
genetic diversity along the genome.

@@

In practice, we do not expect Branch and Site diversity to agree exactly,
because of local differences in mutation rate and intensities of selected mutations.
For instance, regions in which a high proportion of potential mutations are deleterious
are expected to have a lower diversity for two reasons:
first, the deleterious mutations themselves are less likely to be found,
and second, the indirect effect of deleterious mutations on nearby sites reduces typical tree height
and thus diversity at even neutral, linked sites \citep{hudson1994levels,charlesworth1997effects}.
The first effect would cause Site and Branch statistics to differ,
because it effectively reduces the mutation rate in the region
and violates the assumption of independence of mutations given the trees.
However, the second effect does \emph{not} affect the correspondence between Site and Branch statistics,
since it is mediated by tree shape.
It is generally unknown how much mutation rate varies along the genome
\citep[but see][]{supek2015differential}
or how dense targets of selection are~\citep{leffler2012revisiting}.
For this reason, it is very interesting to see how close a correspondence between Site and Branch
statistics it is \emph{possible} to obtain ---
it is tempting to say that the best tree sequence would obtain
as tight a match between Site and Branch statistics as possible,
with remaining variation explained by direct selection and/or mutation rate variation.

@@

%%%%%%%%%%%%%%%%%%%%%
\subsection*{Algorithms}
In this section we describe and analyze the algorithm used
to compute Branch statistics. This is a generalization
of the algorithm used to maintain the number of samples from a
given set that derive from each node in a tree
sequence \cite[Algorithm L]{kelleher2016efficient}. The central
problem shared across Site, Branch and Node statistics
is to efficiently maintain the ``state'' $x_u$ (where $x_u$ is the
sum of the weights for all samples descending
from, and including, node $u$; see Figure~\ref{fig:divergence_diagram}).
As we move along the sequence, the trees change when branches are inserted and removed.
By carefully ordering these insertions and removals,
we ensure that this state can be correctly maintained
using only small adjustments for each tree.
Computing the statistics is then a straightforward matter of applying the summary function $f$
to the node states and aggregating the results appropriately for the
particular statistics mode and windowing options.

@@

\begin{taocpalg}{B}{General Branch statistics}
{Given a list of $n$ sample nodes $S$, corresponding weights $w$ and
summary function $f: \R \rightarrow \R$,
compute the span-normalized statistic $\sigma$
over a tree sequence with sequence length $L$
defined by the node and edge tables $\Nt$ and $\Et$. We assume
that the index vectors $\indexin$ and $\indexout$ have been precomputed.
}

@@

Algorithm \algref{B} begins by setting the initial state for $\pi$, $\beta$,
$x$ and $F$ for each node in the tree sequence, and then sets the values
of $x_u$ and $F_u$ for each of the samples $u$ in $S$.
(The initial state is the ``empty forest'', where no nodes are connected to any others.)
Afterwards, we set our ``running sum'' $s$ and output statistic $\sigma$ to zero, along
with the ``tree left'' variable $t_l$. The $j$ and $k$ variables are
used to keep track of our position in the edge insertion and edge
removal indexes, respectively. After completing initialization in
\algref{B1}, we enter the main tree loop in \algref{B2} which is
run once for each tree in the sequence. As we are processing each
tree, we keep track of the edges that need to be processed with
the $t_l$ variable, which stores the left-most genome coordinate
in the current tree. The first thing we do for a new tree is to
remove any edges that were in the previous tree and are not in the
current tree. These must be the edges in which the right coordinate
is equal to $t_l$, and so \algref{B3} loops over these edges using
the edge removal index $\indexout$. Step \algref{B4} then removes
the branch $u \mapsto v$ corresponding to a single edge, by
subtracting its contribution to the running sum $s$, setting the parent of
$u$ to $-1$ and its branch length to $0$. As shown in
Figure~\ref{fig:algorithm_example}B, removing an edge will result
in the subtree rooted at $u$ becoming disconnected from the rest of the
tree. Step \algref{B5} ensures that the state of the rest of the tree
is correctly maintained by propagating the loss of the state $x_u$
up the tree from $v$. Thus, for each node $v$ that was a
direct ancestor of $u$ in the previous tree, we first subtract
its contribution from the running sum $s$ and remove the contribution
of $x_u$ from its state. Since the state of $v$ has changed, we must
recompute the value of our summary cache $F_v$, and then finally
add the new contribution of $v$ to $s$. (For example, note the changes in $x$
and $F$ for nodes 7 and 8 in Figure~\ref{fig:algorithm_example}B.)

@@

To analyze Algorithm~\algref{B} we will assume that the tree sequence has
been simulated under the standard coalescent with a sample of size
$n$ and population scaled recombination rate $\rho = 4 N_e r$,
where $r$ is the mean number of recombinations per chromosome per unit of time.
Under these conditions, there are $O(n + \rho \log n)$ edges in the tree sequence
\citep{kelleher2016efficient}.  Clearly, each edge is examined
exactly once in steps \algref{B6} and \algref{B7} in order to create
the trees, and at most once in steps \algref{B3} and \algref{B4} (we do
not remove the edges for the final tree). Additionally, each edge that
we insert or remove after the initial building of the tree
may incur the cost of traversing up the tree as
far as the root in steps \algref{B5} and \algref{B8}.
As coalescent genealogies are asymptotically
balanced \citep{li2013coalescent}, the expected number of nodes
on the path to root is $\log_2 n$.
Therefore, the overall running time
of Algorithms~\algref{B} is of order $n + \rho \log(n) \log_2(n)$, which is
\begin{equation}\label{eqn:alg_complexity}
    O\left(n + \rho \log(n)^2 \right).
\end{equation}
Computing a Site statistic will have the same complexity,
as the same internal state is maintained,
and the number of segregating sites is proportional to the number of edges
(with constant equal to the ratio of mutation to recombination rate).
The asymptotic analysis here assumes that we traverse upwards to root
for each edge, but this is not the case. Edges are removed in
oldest-first order, guaranteeing that the deepest node any subtree being
modified is processed first. Therefore, this subtree will be disconnected from the
main tree, ensuring that we traverse upwards all the way to root only once.
Similarly, edges are inserted in youngest-first order, ensuring that subtrees are only
reconnected to the main tree when they are fully built.

@@

To determine how well our theoretical model of time complexity predicts the
running time in \tskit{}'s implementation of the Site statistics algorithm,
we fitted a model based on Equation~\eqref{eqn:alg_complexity}.
\autoref{fig:stats_performance} shows that theoretical predictions match
the observed running time very well. As the overall complexity is
$O(n + \rho \log(n)^2)$, for sufficiently large $n$ the initial term
(representing the cost of building the first tree) will come to dominate.
In this case, we can see this happening at around $n = 10^5$, where there
is a noticeable uptick in the time required per variant. For
longer genomes (i.e., larger values of $\rho$), this cost is amortized over more trees
and is less apparent.

@@

It should be noted here that this example based on simulated data
represents a best-case scenario in terms of the performance advantages
of \tskit{} over the matrix-based methods. For real data, inferred
tree sequences currently contain substantially more edges than we
would expect based on simple neutral theory~\citep{kelleher2019inferring},
and therefore computing statistics will not be as efficient as
for an equivalent simulation. Tree sequence inference
methods are in their infancy, however, and it is likely that
as they improve the number of edges required to encode data will
be reduced. For data simulated natively in tree sequence format
via packages such as \texttt{msprime}~\citep{kelleher2016efficient},
\texttt{SLiM}~\citep{haller2018tree,haller2019slim} and
\texttt{fwdpy11}~\citep{thornton2014c++}, however, the advice is straightforward.
Computing statistics using the algorithms in \tskit{}
will always be more efficient than decoding the genotype
matrix, importing it into another package and computing from the matrix.

@@

% visualization and time decomposition adding a new dimension
Genomic data are naturally distributed on two axes: along the genome, and across geography.
The tree sequence extends this to a third axis: time.
A great many methods in population genetics aim to describe aspects of history,
and accurate (or at least unbiased) tree sequence inference
may shift the focus of these problems from inference to descriptive analysis.
The methods developed here distribute the contribution to various statistics across each tree,
and so could also be used to summarize contributions to various statistics across time.
This could provide, for instance, the time distribution of mutations or branches
contributing positively or negatively to $f_4$ statistics of introgression,
enabling historical interpretations of these signals.
% This additional mode of reporting statistics is an important subject for future work.
% moving past a few summaries in discrete groups to a genome/time/space view of relatedness
The computational toolbox of population genetics
is still mostly composed of statistics designed for analysis of a handful of loci
in a small number of discrete, mostly separated populations.
Both our data and our understanding of the world have moved beyond this.
We hope that the tools developed here will help make it possible
to visualize and analyze genetic variation and genealogical relatedness
along the genome, across space, and through time.
